Bootstrap learning via modular concept discovery

### 2013-04-20 drafts/ijcai2013/ijcai13.tex
Merged changes to ijcai13.tex that were accidentally living in the master branch.

\usepackage{titling}

\setlength{\droptitle}{-1em}   
\date{}

@@

%% \author{Eyal Dechter \\
%% MIT\\
%% USA \\
%% edechter@mit.edu
%% \And
%% Jon Malmaud \\
%% MIT \\ 
%% USA \\
%% malmaud@mit.edu
%% \And 
%% Ryan P.  Adams \\
%% Harvard University\\
%% USA \\
%% rpa@seas.harvard.edu
%% \And
%% Joshua B. Tenenbaum \\
%% MIT\\
%% USA \\
%% jbt@mit.edu
%% }

@@

\maketitle
\begin{abstract}
%\small
 Suppose a learner is faced with a domain of problems about which it
 knows nearly nothing. It does not know the distribution of problems,
 the space of solutions is not smooth, and the reward signal is
 uninformative, providing perhaps a few bits of information but not
 enough to steer it effectively. How can such a learner ever get off
 the ground? A common intuition is that if the solutions to these
 problems share a common structure, and the learner can solve some
 simple problems by brute force, it should be able to extract useful
 components from these solutions and, by composing them, explore the
 solution space more efficiently. Here, we formalize this intuition,
 where the solution space is the space of typed functional programs
 and the gained information is stored as a stochastic grammar over
 programs. We propose an iterative procedure for exploring such
 spaces: in the first step of each iteration, the learner explores a
 finite subset of the domain, guided by a stochastic grammar; in the
 second step, the learner compresses the successful solutions from the
 first step to estimate a new stochastic grammar. We test this
 procedure on symbolic regression and boolean circuit learning and
 show that the learner discovers modular concepts for these
 domains. Whereas the learner is initially unable to solve almost any
 of the posed problems, it quickly acquires the ability to solve many
 problems by gaining abstract knowledge of the structure of the
 solution space.

@@

\section{Related Work}
The idea of discovering and using reusable subcomponents is an old one
in AI. Rendell's 1985 paper~\cite{DBLP:conf/ijcai/Rendell85},
``Substantial Constructive Induction Using Layered Information
Compression: Tractable Feature Formation in Search,'' presented within
a classical search paradigm the idea that compression provides a
heuristic for constructing good representations. Within the Inductive
Logic Programming (ILP) literature, researchers have explored ``Constructive
Induction,'' generating new and reusable logic programming
terms~\cite{DBLP:conf/ijcai/Muggleton87}. We apply these ideas to
learning functional programs, and believe that the modularity and
compositionality afforded by this representation is necessary for
these ideas to be successful.

@@

Our goal is to solve the following minimal feedback multi-task program
induction problem: suppose we are given a set of \textbf{tasks}
$T=\{t_k\}_{k=1}^K$ where each task is a function $t_k : \mathcal{L}
\rightarrow \{0, 1\}$ and where $\mathcal{L}$ is the set of
expressions in our language.  We say that expression $ e \in
\mathcal{L}$ \textbf{solves} $t_k$ if $ t_k(p) = 1$. Our goal is to
solve as many of the tasks in $T$ as we can.

@@

Let $C_e$ be the set of primitive combinators in the leaves of
expressions $e$. We define the probability of an expression ${e \in
  \mathcal{L}}$ to be ${ p(e)~=~\prod_{c \in C_e} p(c | \tau(c)) }$
where $p(c | \tau(c))$ is the probability of using primitive $c$ when
the requesting type is $\tau(c)$. In turn, we define the conditional
probability for each combinator $c_n$ to be $p(c_n| \tau(c_n)) \propto
\frac{p}{\sum_{c_j \in C_{\tau(c_n)}} p_j}$.  That is, $p(c_n | \tau(c_n)$ is
proportional to the probability of sampling $c_n$ from the multinomial
distribution defined by the $p_n$'s, conditioned on the requesting
type, with the constant of proportionality chosen to ensure that the
sum over the probabilities of all expressions converges to a finite
value (so that the distribution over expressions is proper). For
binary trees, this is true whenever the constant of proportionality is
less than~$\frac{1}{4}$\footnote{The probability mass $M_d$ of all
  expressions of depth less than or equal to $d$ can be written with
  the recurrence relation $M_d \leq M_{d-1}^2 + M_1$. This has an
  asymptotic fixed point, as $d \rightarrow \infty$ if $x = x^2 + M_1$
  has a solution, which is true if $M_1~\leq~1/4$.}.

@@

\subsection{Best-first enumeration of programs}
In order to explore the frontier of most promising programs, we need a
procedure that enumerates the $N$ most probable expressions. There has
been recent interest in this problem, most of it focused on
enumerating the shortest program satisfying some
criterion~\cite{DBLP:conf/sfp/Katayama05}
\cite{DBLP:conf/aaip/YakushevJ09}. Our approach is to formulate the
problem as a best-first exploration of an AND/OR tree
(Figure~\ref{fig:andor})~\cite{nilsson1982principles}~\cite{DBLP:journals/cacm/Hall73}. In
words, the enumeration procedure is best described by the following
recursion: every program is either a primitive combinator or it is a
left child program applied to a right child program.

@@

Our best-first policy is to expand the next partial solution with the
highest value. If a partial solution has multiple leaves that are not
primitive combinators, we expand the left most one first, since it is
the choice of left combinator that defines the type of its right
combinator in an application pair. That is, the parent node applies
its constraint on valid solutions via the left child, not the right
child. We want to apply constraints as early as possible, so we expand
the left child first to prune more partial solutions.

>>>>

@@

\author{Eyal Dechter \\
MIT\\
USA \\
edechter@mit.edu
\And
Jon Malmaud \\
MIT \\ 
USA \\
malmaud@mit.edu
\And 
Ryan P.  Adams \\
Harvard University\\
USA \\
rpa@seas.harvard.edu
\And
Joshua B. Tenenbaum \\
MIT\\
USA \\
jbt@mit.edu
}

@@

\maketitle
\begin{abstract}
%\small
 Suppose a learner is faced with a domain of problems about which it
 knows nearly nothing. It does not know the distribution of problems,
 the space of solutions is not smooth, and the reward signal is
 uninformative, providing perhaps a few bits of information but not
 enough to steer it effectively. How can such a learner ever get off
 the ground? A common intuition is that if the solutions to these
 problems share a common structure, and the learner can solve some
 simple problems by brute force, it should be able to extract useful
 components from these solutions and, by composing them, explore the
 solution space more efficiently. Here, we formalize this intuition,
 where the solution space is the space of typed functional programs
 and the gained information is stored as a stochastic grammar over
 programs. We propose an iterative procedure for exploring such
 spaces: in the first step of each iteration, the learner explores a
 finite subset of the domain, guided by a stochastic grammar; in the
 second step, the learner compresses the successful solutions from the
 first step to estimate a new stochastic grammar. We test this
 procedure on symbolic regression and boolean circuit learning and
 show that the learner discovers modular concepts for these
 domains. Whereas the learner is able to solve almost none of the
 posed problems in the procedure's first iteration, it rapidly becomes
 able to solve a large number by gaining abstract knowledge of the
 structure of the solution space.

@@

\section{Related Work}
The idea of discovering and using reusable subcomponents is part of an
old tradition in AI. For instance, it is one of the central themes in
Herb Simon's 1969 ``The Sciences of the
Artificial''~\cite{simon1996sciences}. Rendell's 1985
paper~\cite{DBLP:conf/ijcai/Rendell85}, ``Substantial Constructive
Induction Using Layered Information Compression: Tractable Feature
Formation in Search,'' presented within a classical search paradigm
the idea that compression provides a heuristic for constructing good
representations. Within the Inductive Logic Programming (ILP)
literature, researchers have explored ``Constructive Induction,''
generating new and reusable logic programming
terms~\cite{DBLP:conf/ijcai/Muggleton87}. Later, predicate invention
was also applied in the multitask
setting~\cite{DBLP:conf/ilp/KhanMP98}. We apply these ideas to
learning functional programs, and believe that the modularity and
compositionality afforded by this representation is necessary for
these ideas to be successful.

@@

Our goal is to solve the following minimal feedback multi-task program
induction problem: suppose we are given a set of \textbf{tasks}
$T=\{t_k\}_{k=1}^K$ where each task is a function $t_k : \mathcal{L}
\rightarrow \{0, 1\}$ and where $\mathcal{L}$ is the set of
expressions in our language.  We say that expression $ e \in
\mathcal{L}$ \textbf{solves} $t_k$ if $ t_k(e) = 1$. Our goal is to
solve as many of the tasks in $T$ as we can.

@@

Let $C_e$ be the set of primitive combinators in the leaves of
expressions $e$. We define the probability of an expression ${e \in
  \mathcal{L}}$ to be ${ p(e)~=~\prod_{c \in C_e} p(c | \tau(c)) }$
where $p(c | \tau(c))$ is the probability of using primitive $c$ when
the requesting type is $\tau(c)$. In turn, we define the conditional
probability for each combinator $c_n$ to be $p(c_n| \tau(c_n)) \propto
\frac{p_n}{\sum_{c_j \in C_{\tau(c_n)}} p_j}$.  That is, $p(c_n | \tau(c_n)$ is
proportional to the probability of sampling $c_n$ from the multinomial
distribution defined by the $p_n$'s, conditioned on the requesting
type, with the constant of proportionality chosen to ensure that the
sum over the probabilities of all expressions converges to a finite
value (so that the distribution over expressions is proper). For
binary trees, this is true whenever the constant of proportionality is
less than~$\frac{1}{4}$\footnote{The probability mass $M_d$ of all
  expressions of depth less than or equal to $d$ can be written with
  the recurrence relation $M_d \leq M_{d-1}^2 + M_1$. This has an
  asymptotic fixed point, as $d \rightarrow \infty$ if $x = x^2 + M_1$
  has a solution, which is true if $M_1~\leq~1/4$.}.

@@

\subsection{Best-first enumeration of programs}
In order to explore the frontier of most promising programs, we need a
procedure that enumerates the $N$ most probable expressions. There has
been recent interest in this problem, most of it focused on
enumerating the shortest program satisfying some
criterion~\cite{DBLP:conf/sfp/Katayama05}
\cite{DBLP:conf/aaip/YakushevJ09}. Our approach is to formulate the
problem as a best-first exploration of an AND/OR tree
(Figure~\ref{fig:andor})~\cite{nilsson1982principles}~\cite{DBLP:journals/cacm/Hall73}. In
words, the enumeration procedure is best described by the following
recursion: every program is either a primitive combinator OR it is a
left child program applied to a right child program (that is, a left
child AND a right child).

@@

Our best-first policy is to expand the next partial solution
with the highest value. If a partial solution has multiple leaves that
are not primitive combinators, we expand the left most one first,
since it is the choice of left combinator that defines the type of its
right combinator in an application pair. That is, the parent node
applies its constraint on valid solutions via the left child, not the
right child. We want to apply constraints as early as possible, so
expanding left child first prunes more partial solutions.
